---
title: Using the DuckDB R API
author: Ned Letcher
output: html_document
---

# Using the DuckDB R API

## Technical requirements 

In order to run the examples in this notebook, you'll need to install the R dependencies for this project. You can do this by running the following command in your R session. 

    install.packages(c("duckdb", "tidyverse")) 

For complete instructions on how to set up your environment for working through the examples, please consult the *Technical requirements* section of this chapter in the book.

## Working with DuckDB using R’s DBI 

```{r}
library("DBI")
```

### Connecting to DuckDB

```{r}
disk_conn <- dbConnect(duckdb::duckdb(), dbdir = "quack.duckdb")
```

```{r}
read_only_conn <- dbConnect(duckdb::duckdb(), dbdir = "quack.duckdb", read_only = TRUE)
```

```{r}
mem_conn <- dbConnect(duckdb::duckdb(), dbdir = ":memory:")
```

### Reading and writing tables

```{r}
library(tidyverse) 

dogs_df <- read_delim(
    file = "NYC_Dog_Licensing_Dataset.csv",
    delim = ";",
    col_types = cols(
        LicenseIssuedDate = col_date("%m/%d/%Y"),
        LicenseExpiredDate = col_date("%m/%d/%Y"),
        AnimalGender = col_factor(levels = c("M", "F")),
    ),
)

head(dogs_df, 5)
```

```{r}
conn <- dbConnect(duckdb::duckdb(), dbdir = "nyc_dogs.duckdb")

dbWriteTable(conn, "dogs_from_df", dogs_df)
```

```{r}
dbListTables(conn)
```

```{r}
dbListFields(conn, "dogs_from_df")
```

```{r}
dogs_from_duckdb_df <- dbReadTable(conn, "dogs_from_df")

tail(dogs_from_duckdb_df, 5)
```

```{r}
dbCreateTable(conn, "dogs", dogs_df) 

dbReadTable(conn, "dogs") 
```

```{r}
dbCreateTable( 
    conn,
    "good_dogs",
    c(name = "character", birthday = "date", age = "double")
) 

dbListTables(conn)
```

```{r}
dbDataType(conn, "Monty")

dbDataType(conn, 6)

dbDataType(conn, today())
```

```{r}
dbRemoveTable(conn, "good_dogs") 

dbListTables(conn)
```

### Querying and executing SQL statements

```{r}
dbGetQuery(
    conn,
    "
    SELECT
        ZipCode,
        count(*) AS num_registrations
    FROM dogs_from_df
    GROUP BY ZipCode
    ORDER BY num_registrations DESC
    LIMIT 10
    "
)
```

```{r}
dbExecute(
    conn,
    " 
    CREATE OR REPLACE VIEW nyc_dogs_csv AS
    SELECT *
    FROM 'NYC_Dog_Licensing_Dataset.csv'
    "
)
```

```{r}
dbGetQuery(
    conn,
    " 
    SELECT *
    FROM nyc_dogs_csv
    USING SAMPLE 3
    "
)
```

```{r}
dbExecute(
    conn,
    " 
    CREATE OR REPLACE TABLE nyc_dogs AS
    SELECT *
    FROM nyc_dogs_csv
    "
)

dbListTables(conn)
```

```{r}
dogs_df1 = dbGetQuery(conn, "SELECT * FROM nyc_dogs")

dogs_df2 = dbReadTable(conn, "nyc_dogs")
```

### Using Prepared statements  

```{r}
dbGetQuery(
    conn,
    "
    SELECT *
    FROM nyc_dogs
    WHERE BreedName = ? AND AnimalGender = ?
    LIMIT ?
    ",
    list("Beagle", "M", 3)
)
```

```{r}
result <- dbSendStatement(conn, "INSERT INTO nyc_dogs VALUES (?, ?, ?, ?, ?, ?, ?, ?)")

dbBind(result, list("John", "M", 2013, "Jack Russell Terrier", 10261, "2014-07-12", "2017-08-09", 2016))

dbBind(result, list("Lady Fluffina", "F", 2014, "Bichon Frisé", 10302, "2014-08-22", "2017-09-25", 2016))

dbClearResult(result)
```

```{r}
tail(dbReadTable(conn, "nyc_dogs"), 2)
```

### Disconnecting from DuckDB  

```{r}
dbDisconnect(conn, shutdown = TRUE)
```

```{r}
run_query <- function(db_path, sql_query) {
    conn <- dbConnect(duckdb::duckdb(), dbdir = db_path)
    on.exit(dbDisconnect(conn, shutdown = TRUE), add = TRUE)
    dbGetQuery(conn, sql_query)
}
```

```{r}
run_query("nyc_dogs.duckdb", "SELECT * FROM nyc_dogs USING SAMPLE 3")
```

## Registering R objects as virtual tables 

### Registering a dataframe as a virtual table

```{r}
conn <- dbConnect(duckdb::duckdb(), dbdir = "nyc_dogs.duckdb")

duckdb::duckdb_register(conn, "dogs_df_view", dogs_df)

dbListTables(conn)
```

```{r}
dbGetQuery(conn, "SELECT * FROM dogs_df_view USING SAMPLE 3")
```

```{r}
duckdb::duckdb_unregister(conn, "dogs_df_view")

dbListTables(conn)
```

### Registering an Arrow table as a virtual table

```{r}
library(arrow) 

dogs_arrow = arrow::arrow_table(dogs_df)
```

```{r}
duckdb::duckdb_register_arrow(conn, "dogs_arrow_view", dogs_arrow)
```

```{r}
dbGetQuery(conn, "SELECT * FROM dogs_arrow_view USING SAMPLE 3")
```

```{r}
duckdb::duckdb_list_arrow(conn)
```

```{r}
duckdb::duckdb_unregister_arrow(conn, "dogs_arrow_view")
```

## Using DuckDB with `dplyr`

### Using `dplyr` to query dataframes

```{r}
dbReadTable(conn, "nyc_dogs_csv") |>
    group_by(year=year(LicenseIssuedDate)) |>
    summarise(count = n()) |>
    arrange(desc(count))
```

```{r}
dbReadTable(conn, "nyc_dogs_csv") |>
    count(year=year(LicenseIssuedDate), name = "count", sort = TRUE)
```

### Using `dplyr` to query DuckDB tables via `dbplyr`

```{r}
tbl(conn, "nyc_dogs_csv")
```

```{r}
tbl(conn, "nyc_dogs_csv") |>
    head(3) |> 
    collect()
```

```{r}
tbl(conn, "nyc_dogs_csv") |>
    head(3) |>
    show_query()
```

### Data wrangling with `dplyr`

```{r}
unique_dogs <- tbl(conn, "nyc_dogs_csv") |>
    filter(!AnimalName %in% c("UNKNOWN", "NAME NOT PROVIDED", "NAME", "NONE")) |>
    filter(!BreedName %in% c("Unknown", "Not Provided")) |>
    distinct(AnimalName, AnimalGender, AnimalBirthYear, BreedName, ZipCode)
```

```{r}
unique_dogs |>
    head(5) |>
    collect()
```

```{r}
table_query <- unique_dogs |>
    filter(AnimalGender == "M") |>
    count(AnimalName, name = "num_dogs", sort = TRUE) |>
    head(10)
```

```{r}
table_query |>
    show_query()
```

```{r}
table_query |>
    collect()
```

```{r}
pop_dog_names <- unique_dogs |>
    filter(AnimalGender == "F", AnimalBirthYear > 2010) |>
    count(AnimalBirthYear, AnimalName, name = "NumDogs") |>
    slice_max(by = AnimalBirthYear, order_by = NumDogs) |>
    arrange(AnimalBirthYear)
```

```{r}
pop_dog_names |>
    collect()
```

```{r}
issued_by_year = tbl(conn, "nyc_dogs_csv") |>
    count(
        LicenseIssuedYear=year(LicenseIssuedDate),
        name = "Count",
        sort = TRUE
    )

issued_by_year |>
    collect()
```

```{r}
issued_by_year |> 
    show_query()
```

```{r}
unique_dogs |>
    mutate(Hash=md5(concat(AnimalName, ZipCode))) |>
    head(3) |>
    collect()
```

```{r}
unique_dogs |>
    count(AnimalName, name = "Count") |>
    filter(Count >= 100) |>
    mutate(EditDistance=round(jaro_winkler_similarity(AnimalName, "BELLA"), 3)) |>
    arrange(desc(EditDistance)) |>
    head(10) |>
    collect()
```

## Summary

